---
title: "Statistical Inference Course Project"
author: "Rui Mendes"
date: "24-01-2015"
output: 
    tufterhandout::html_tufte_handout
---

### Part1: "Simulation Exercises"

The exponential distribution can be simulated in R with `rexp(n, lambda)` where
`lambda` $\lambda$ is the rate parameter. The mean of exponential distribution is 
$1/\lambda$ and the standard deviation is also $1/\lambda$. In this project,
we set $\lambda=0.2$ for all simulations. 

In this simulation, we will investigate the distribution of
averages of 40 exponentials.

Let's do a thousand simulated averages of 40 exponentials.

```{r}
## Define seed (to make reproducible)
set.seed(3)

## Set the value of lambda
lambda <- 0.2

## Setnumber of simulations
num_sim <- 1000

## Set the sample size
sample_size <- 40

## Define a matrix with the sample
sim <- matrix(rexp(num_sim*sample_size, rate=lambda), num_sim, sample_size)
row_means <- rowMeans(sim)
```

The distribution of sample means is as follows.

```{r echo=FALSE}
## Plot the histogram of averages
hist(row_means, breaks=50, prob=TRUE,
     main="Distribution of averages of samples,
     drawn from exponential distribution with lambda=0.2",
     xlab="", ylab="Density", col="grey")

## Add the density of the averages of samples
lines(density(row_means), col="green")

## Theoretical center of distribution
abline(v=1/lambda, col="red")

## Theoretical density of the averages of samples
xfit <- seq(min(row_means), max(row_means), length=100)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(sample_size)))
lines(xfit, yfit, pch=22, col="red", lty=2)

## Add the legend
legend('topright', c("Simulation", "Theoretical"), lty=c(1,2), col=c("green", "red"))
```

The distribution of sample means is centered at `r mean(row_means)`
and the theoretical center of the distribution is $\lambda^{-1}$ = `r 1/lambda`.

The variance of sample means is `r var(row_means)` where the theoretical variance
of the distribution is $\sigma^2 / n = 1/(\lambda^2 n) = 1/(0.04 \times 40)$ =
`r 1/(0.04 * 40)`.

Due of the central limit theorem, the averages of samples follow the normal
distribution. 

The figure plotted above also shows the density computed using the histogram and the
normal density plotted with theoretical mean and variance values. Also, the
q-q plot below suggests the normal distribution

```{r echo=FALSE}
qqnorm(row_means); qqline(row_means)
```

Finally, let's see the evaluation of the coverage for the confidence interval with
$1/\lambda = \bar{X} \pm 1.96 \frac{S}{\sqrt{n}}$

```{r echo=FALSE}
## Define lambda values
lambda_vals <- seq(4, 6, by=0.01)

## Construct coverage
coverage <- sapply(lambda_vals, function(lamb) {
    mu_hats <- rowMeans(matrix(rexp(sample_size*num_sim, rate=0.2),
                               num_sim, sample_size))
    ll <- mu_hats - qnorm(0.975) * sqrt(1/lambda**2/sample_size)
    ul <- mu_hats + qnorm(0.975) * sqrt(1/lambda**2/sample_size)
    mean(ll < lamb & ul > lamb)
})

library(ggplot2)
## Plot the graphic
qplot(lambda_vals, coverage) + geom_hline(yintercept=0.95)
```

The 95% confidence intervals for the rate parameter ($\lambda$) to be estimated
($\hat{\lambda}$) are
$\hat{\lambda}_{low} = \hat{\lambda}(1 - \frac{1.96}{\sqrt{n}})$ agnd
$\hat{\lambda}_{upp} = \hat{\lambda}(1 + \frac{1.96}{\sqrt{n}})$.
As can be seen from the plot above, for selection of $\hat{\lambda}$ around 5,
the average of the sample mean falls within the confidence interval at least 95% of the time.
Note that the true rate, $\lambda$ is 5.

This file can be found [here](https://github.com/rui-mendes/StatisticalInference/tree/master/Part1) and/or [here](http://rpubs.com/ruimendes/statistical-inference-proj-part1).